---
title: "SNP Direct and Indirect Effects"
author: "James Long"
date: "10/1/2018"
output: html_document
header-includes:
  - \usepackage{color}
---

<style>
.column-left{
  float: left;
  width: 50%;
  text-align: center;
}
.column-right{
  float: right;
  width: 50%;
  text-align: center;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(knitr.kable.NA = '')
options(scipen = 999)
library(graph)
library(kableExtra)
devtools::load_all("../mediateR/")
library(mediateR)
set.seed(280920181)
```
\newcommand{\Var}{\text{Var}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\argmin}[1]{\underset{#1}{\operatorname{argmin}}\text{ }}
\newcommand{\argmax}[1]{\underset{#1}{\operatorname{argmax}}}
\newcommand{\ind}[1]{\mathbf{1}_{#1}}
\newcommand{\todo}[1]{\color{red}{\text{TODO: #1}}}

## Simulation 1: Linear, Low Dimensional

```{r simulation_params,echo=FALSE,eval=TRUE}
n <- 500
sim_params <- QuickSim(n,3,2,"gaussian")
dat <- SimulateData(sim_params)
```

We simulate `r n` observations from the "True Graph" using independent, binary SNPs drawn from Bernoulli(1/2) and path coefficients given below.

```{r plot_make,echo=FALSE,eval=TRUE}
gR <- MakeGraphNELObject(sim_params$path,sim_params$xx_direct,sim_params$mm_direct)
attrs <- list()
attrs$edge <- list()
attrs$edge$fontsize <- 12
edgeAttrs <- MakeedgeAttrs(sim_params$path_model,sim_params$xx_direct,sim_params$mm_direct)
```

```{r compute_estimate,echo=FALSE,eval=TRUE}
fit <- ComputePath(dat)
eff_est <- ComputeEffectsLinear(fit)
```


```{r plot_subgraph,echo=FALSE,eval=TRUE,fig.height=6,fig.width=12}
gR2 <- MakeGraphNELObject(fit$path_model,fit$xx_direct,fit$mm_direct)
attrs2 <- list()
attrs2$edge <- list()
attrs2$edge$fontsize <- 12
edgeAttrs2 <- MakeedgeAttrs(fit$path_model,fit$xx_direct,fit$mm_direct)
par(mfcol=c(1,2))
plot(gR,edgeAttrs=edgeAttrs,attrs=attrs,main="True Graph")
plot(gR2,edgeAttrs=edgeAttrs2,attrs=attrs2,main="Estimated Graph")
```


```{r merge_true_est,echo=FALSE,eval=TRUE}
eff <- ComputeEffectsLinear(sim_params)
eff_comb <- matrix(0,nrow(eff),ncol=2*ncol(eff))
for(ii in 1:ncol(eff)){
  eff_comb[,2*ii-1] <- eff_est[,ii]
  eff_comb[,2*ii] <- eff[,ii]
}
for(ii in (ncol(dat$xx)+1):(ncol(dat$xx)+ncol(dat$mm))){
  eff_comb[ii,3:ncol(eff_comb)] <- NA
}
colnames(eff_comb) <- rep(c("est","true"),ncol(eff))
rownames(eff_comb) <- rownames(eff)
```



```{r bootstrap,echo=FALSE,eval=TRUE}
B <- 100
eff_est_boot <- vector("list",length=B)
for(ii in 1:B){
  ix <- sample(1:sim_params$n,replace=TRUE)
  dat_sub <- SubsetDat(dat,ix)
  fit <- ComputePath(dat_sub)
  eff_est_boot[[ii]] <- ComputeEffectsLinear(fit)
}
```

<div class = "row">
<div class="column-left">
```{r kable_table,echo=FALSE}
kable(eff_comb,digits=2) %>%
  kable_styling(bootstrap_options = "striped", full_width = F) %>%
  add_header_above(c(" ", "direct" = 2, "indirect" = 2, "total" = 2))
```
</div>
<div class="column-right">
```{r bootstrap_plot,echo=FALSE,fig.height=4.5,fig.width=4.5}
total_xx1 <- vapply(eff_est_boot,function(x){x[1,3]},c(0))
hist(total_xx1,main="Bootstrap Samples SNP1 Total Effect",xlab="Estimate")
abline(v=eff[1,3],lwd=3,col='red')
abline(v=eff_est[1,3],lwd=3,col='blue')
legend("topright",c("Truth","Point Estimate"),col=c("red","blue"),lwd=2)
```
</div>
</div>



## Simulation 2: Logistic, Low Dimensional

Same as simulation 1 but with binomial link function connecting y with snps and gene sets.

```{r simulation_params2,echo=FALSE,eval=TRUE}
sim_params$family <- "binomial"
dat <- SimulateData(sim_params)
```

```{r plot_make2,echo=FALSE,eval=TRUE}
gR <- MakeGraphNELObject(sim_params$path,sim_params$xx_direct,sim_params$mm_direct)
attrs <- list()
attrs$edge <- list()
attrs$edge$fontsize <- 12
edgeAttrs <- MakeedgeAttrs(sim_params$path_model,sim_params$xx_direct,sim_params$mm_direct)
```

```{r compute_estimate2,echo=FALSE,eval=TRUE}
fit <- ComputePath(dat)
```

```{r plot_subgraph2,echo=FALSE,eval=TRUE,fig.height=6,fig.width=12}
gR2 <- MakeGraphNELObject(fit$path_model,fit$xx_direct,fit$mm_direct)
attrs2 <- list()
attrs2$edge <- list()
attrs2$edge$fontsize <- 12
edgeAttrs2 <- MakeedgeAttrs(fit$path_model,fit$xx_direct,fit$mm_direct)
par(mfcol=c(1,2))
plot(gR,edgeAttrs=edgeAttrs,attrs=attrs,main="True Graph")
plot(gR2,edgeAttrs=edgeAttrs2,attrs=attrs2,main="Estimated Graph")
```

```{r xx_effects_decomp2, echo=FALSE,eval=TRUE}
## compute true effects by approximating integral with large sample size
sim_params2 <- sim_params
sim_params2$n <- 1e6 ## use million observations to approximate effects
dat2 <- SimulateData(sim_params2)
direct_t <- ComputeEffectxx(dat2,sim_params,"direct")
indirect_t <- ComputeEffectxx(dat2,sim_params,"indirect")
total_t <- ComputeEffectxx(dat2,sim_params,"total")

direct <- ComputeEffectxx(dat,fit,"direct")
indirect <- ComputeEffectxx(dat,fit,"indirect")
total <- ComputeEffectxx(dat,fit,"total")
eff_comb <- cbind(direct,direct_t,indirect,indirect_t,total,total_t)
rownames(eff_comb) <- names(fit$xx_direct)
colnames(eff_comb) <- c("est","true","est","true","est","true")
xx1total_sample <- eff_comb[1,5]
xx1total_true <- eff_comb[1,6]
```

```{r bootstrap2,echo=FALSE,eval=TRUE}
B <- 100
total_xx1 <- rep(0,length=B)
for(ii in 1:B){
  ix <- sample(1:sim_params$n,replace=TRUE)
  dat_boot <- SubsetDat(dat,ix)
  fit_boot <- ComputePath(dat_boot)
  total_xx1[ii] <- ComputeEffectxx(dat_boot,fit_boot,"total")[1]
}
```


<div class = "row">
<div class="column-left">
```{r kable_table2,echo=FALSE}
kable(eff_comb,digits=2) %>%
  kable_styling(bootstrap_options = "striped", full_width = F) %>%
  add_header_above(c(" ", "direct" = 2, "indirect" = 2, "total" = 2))
```
</div>
<div class="column-right">
```{r bootstrap_plot2,echo=FALSE,fig.height=4.5,fig.width=4.5}
hist(total_xx1,main="Bootstrap Samples SNP1 Total Effect",xlab="Estimate")
abline(v=xx1total_sample,lwd=3,col='blue')
abline(v=xx1total_true,lwd=3,col='red')
legend("topright",c("Point Estimate","True"),col=c("blue","red"),lwd=2)
```
</div>
</div>












## Simulation 3: Survival, Low Dimensional

Same as simulation 1 and 3 but with exponentially distributed survival times. The rate is $e^{\beta^Tx}$ where $x$ are the covariates. This model can be fit with Cox Proportional Hazards.

```{r simulation_params3,echo=FALSE,eval=TRUE}
sim_params$family <- "cox"
dat <- SimulateData(sim_params)
rmean <- max(as.matrix(dat$y)[,1])
```


Same as simulation 1 and 3 but with exponentially distributed survival times. The rate is $e^{\beta^Tx}$ where $x$ are the covariates. This model can be fit with Cox Proportional Hazards. We compute the effects on mean survival restricted to `r round(rmean)`.

```{r plot_make3,echo=FALSE,eval=TRUE}
gR <- MakeGraphNELObject(sim_params$path,sim_params$xx_direct,sim_params$mm_direct)
attrs <- list()
attrs$edge <- list()
attrs$edge$fontsize <- 12
edgeAttrs <- MakeedgeAttrs(sim_params$path_model,sim_params$xx_direct,sim_params$mm_direct)
```

```{r compute_estimate3,echo=FALSE,eval=TRUE}
fit <- ComputePath(dat)
```

```{r plot_subgraph3,echo=FALSE,eval=TRUE,fig.height=6,fig.width=12}
gR2 <- MakeGraphNELObject(fit$path_model,fit$xx_direct,fit$mm_direct)
attrs2 <- list()
attrs2$edge <- list()
attrs2$edge$fontsize <- 12
edgeAttrs2 <- MakeedgeAttrs(fit$path_model,fit$xx_direct,fit$mm_direct)
par(mfcol=c(1,2))
plot(gR,edgeAttrs=edgeAttrs,attrs=attrs,main="True Graph")
plot(gR2,edgeAttrs=edgeAttrs2,attrs=attrs2,main="Estimated Graph")
```

```{r xx_effects_decomp3, echo=FALSE,eval=TRUE}
## compute true effects by approximating integral with large sample size
## use the maximum observed time as the restricted mean time
sim_params2 <- sim_params
sim_params2$n <- 1e4 ## use many observations to approximate effects
dat2 <- SimulateData(sim_params2)
fit2 <- ComputePath(dat2) ## perhaps better to modify fit by entering new coefficients

direct_t <- ComputeEffectxx(dat2,fit2,"direct",rmean=rmean)
indirect_t <- ComputeEffectxx(dat2,fit2,"indirect",rmean=rmean)
total_t <- ComputeEffectxx(dat2,fit2,"total",rmean=rmean)

direct <- ComputeEffectxx(dat,fit,"direct",rmean=rmean)
indirect <- ComputeEffectxx(dat,fit,"indirect",rmean=rmean)
total <- ComputeEffectxx(dat,fit,"total",rmean=rmean)
eff_comb <- cbind(direct,direct_t,indirect,indirect_t,total,total_t)
rownames(eff_comb) <- names(fit$xx_direct)
colnames(eff_comb) <- c("est","true","est","true","est","true")
xx1total_sample <- eff_comb[1,5]
xx1total_true <- eff_comb[1,6]
```

```{r bootstrap3,echo=FALSE,eval=TRUE}
B <- 100
total_xx1 <- rep(0,length=B)
for(ii in 1:B){
  ix <- sample(1:sim_params$n,replace=TRUE)
  dat_boot <- SubsetDat(dat,ix)
  fit_boot <- ComputePath(dat_boot)
  total_xx1[ii] <- ComputeEffectxx(dat_boot,fit_boot,"total",rmean=rmean)[1]
}
```


<div class = "row">
<div class="column-left">
```{r kable_table3,echo=FALSE}
kable(eff_comb,digits=2) %>%
  kable_styling(bootstrap_options = "striped", full_width = F) %>%
  add_header_above(c(" ", "direct" = 2, "indirect" = 2, "total" = 2))
```
</div>
<div class="column-right">
```{r bootstrap_plot3,echo=FALSE,fig.height=4.5,fig.width=4.5}
hist(total_xx1,main="Bootstrap Samples SNP1 Total Effect",xlab="Estimate")
abline(v=xx1total_sample,lwd=3,col='blue')
abline(v=xx1total_true,lwd=3,col='red')
legend("topright",c("Point Estimate","True"),col=c("blue","red"),lwd=2)
```
</div>
</div>

