---
title: "High Dimensional Causal Effects"
author: "James Long"
date: "10/29/2018"
output: html_document
header-includes:
  - \usepackage{color}
---

<style>
.column-left{
  float: left;
  width: 50%;
  text-align: center;
}
.column-right{
  float: right;
  width: 50%;
  text-align: center;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(knitr.kable.NA = '')
options(scipen = 999)
library(graph)
library(kableExtra)
devtools::load_all("../mediateR/")
set.seed(28092018)
```
\newcommand{\Var}{\text{Var}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\argmin}[1]{\underset{#1}{\operatorname{argmin}}\text{ }}
\newcommand{\argmax}[1]{\underset{#1}{\operatorname{argmax}}}
\newcommand{\ind}[1]{\mathbf{1}_{#1}}
\newcommand{\todo}[1]{\color{red}{\text{TODO: #1}}}


## Simulation 1: Highish Dimensional, Linear

```{r simulation_params_high,echo=FALSE,eval=TRUE}
n <- 1000
nxx <- 5
nmm <- 100
sim_params <- QuickSim(n,nxx,nmm,"gaussian")
dat <- SimulateData(sim_params)
```

We simulate `r n` observations from `r nxx` SNPs and `r nmm` gene sets.

The **relevant subgraph** is the set of SNPs and gene sets which are connected to $y$ either directly or through gene sets (in the case of SNPs). To keep the number of nodes and connections reasonable, we only plot the relevant subgraph.

The path structure and direct effects for SNP1,SNP2,SNP3,gs1,and gs2 are fixed. For SNPs 4 and up and gs 3 and up, SNPs are connected to each gene sets with probability $0.1$ and the path coefficients are randomly either -1/2 or 1/2.

The Elastic net ($L_1$ + $L_2$ penalty) to estimate parameters with cross validation used to select tuning parameters.


The true and estimated relevant subgraphs are:

```{r plot_make,echo=FALSE,eval=TRUE}
path_sub <- FindSubgraph(sim_params)
gR <- MakeGraphNELObject(FindSubgraph(sim_params),
                         sim_params$xx_direct[rownames(sim_params$path) %in% rownames(path_sub)],
                         sim_params$mm_direct[colnames(sim_params$path) %in% colnames(path_sub)])
attrs <- list()
attrs$edge <- list()
attrs$edge$fontsize <- 12
edgeAttrs <- MakeedgeAttrs(sim_params$path_model,sim_params$xx_direct,sim_params$mm_direct)
```

```{r compute_estimate,echo=FALSE,eval=TRUE}
fit <- ComputePath(dat,reg=TRUE)
eff_est <- ComputeEffectsLinear(fit)
```

```{r plot_subgraph,echo=FALSE,eval=TRUE,fig.height=6,fig.width=12}
fit_sub <- fit
path_sub <- FindSubgraph(fit_sub)
fit_sub$path_model <- path_sub
fit_sub$xx_direct <- fit_sub$xx_direct[rownames(fit_sub$path_model)]
fit_sub$mm_direct <- fit_sub$mm_direct[colnames(fit_sub$path_model)]

gR2 <- MakeGraphNELObject(fit_sub$path_model,fit_sub$xx_direct,fit_sub$mm_direct)
attrs2 <- list()
attrs2$edge <- list()
attrs2$edge$fontsize <- 12
edgeAttrs2 <- MakeedgeAttrs(fit_sub$path_model,fit_sub$xx_direct,fit_sub$mm_direct)
par(mfcol=c(1,2))
plot(gR,edgeAttrs=edgeAttrs,attrs=attrs,main="True Relevant Subgraph")
plot(gR2,edgeAttrs=edgeAttrs2,attrs=attrs2,main="Estimated Relevant Subgraph")
```


```{r merge_true_est,echo=FALSE,eval=TRUE}
eff <- ComputeEffectsLinear(sim_params)
eff_comb <- matrix(0,nrow(eff),ncol=2*ncol(eff))
for(ii in 1:ncol(eff)){
  eff_comb[,2*ii-1] <- eff_est[,ii]
  eff_comb[,2*ii] <- eff[,ii]
}
for(ii in (ncol(dat$xx)+1):(ncol(dat$xx)+ncol(dat$mm))){
  eff_comb[ii,3:ncol(eff_comb)] <- NA
}
colnames(eff_comb) <- rep(c("est","true"),ncol(eff))
rownames(eff_comb) <- rownames(eff)
eff_comb <- FindInterestingEffs(eff_comb)
```





<div class = "row">
<div class="column-left">
```{r kable_table,echo=FALSE}
kable(eff_comb,digits=3) %>%
  kable_styling(bootstrap_options = "striped", full_width = F) %>%
  add_header_above(c(" ", "direct" = 2, "indirect" = 2, "total" = 2))
```
</div>
<div class="column-right">
</div>
</div>





## Simulation 2: Highish Dimensional, Logistic Model

Same setup as before but with a logistic model linking y and the snps and gene sets.

```{r simulation_params2,echo=FALSE,eval=TRUE}
sim_params$family <- "binomial"
dat <- SimulateData(sim_params)
```

```{r plot_make2,echo=FALSE,eval=TRUE}
path_sub <- FindSubgraph(sim_params)
gR <- MakeGraphNELObject(FindSubgraph(sim_params),
                         sim_params$xx_direct[rownames(sim_params$path) %in% rownames(path_sub)],
                         sim_params$mm_direct[colnames(sim_params$path) %in% colnames(path_sub)])
attrs <- list()
attrs$edge <- list()
attrs$edge$fontsize <- 12
edgeAttrs <- MakeedgeAttrs(sim_params$path_model,sim_params$xx_direct,sim_params$mm_direct)
```

```{r compute_estimate2,echo=FALSE,eval=TRUE}
fit <- ComputePath(dat,reg=TRUE)
```

```{r plot_subgraph2,echo=FALSE,eval=TRUE,fig.height=6,fig.width=12}
path_sub <- FindSubgraph(fit)
fit_plot <- fit
fit_plot$path_model <- path_sub
fit_plot$xx_direct <- fit_plot$xx_direct[rownames(fit_plot$path_model)]
fit_plot$mm_direct <- fit_plot$mm_direct[colnames(fit_plot$path_model)]

gR2 <- MakeGraphNELObject(fit_plot$path_model,fit_plot$xx_direct,fit_plot$mm_direct)
attrs2 <- list()
attrs2$edge <- list()
attrs2$edge$fontsize <- 12
edgeAttrs2 <- MakeedgeAttrs(fit_plot$path_model,fit_plot$xx_direct,fit_plot$mm_direct)
par(mfcol=c(1,2))
plot(gR,edgeAttrs=edgeAttrs,attrs=attrs,main="True Subgraph")
plot(gR2,edgeAttrs=edgeAttrs2,attrs=attrs2,main="Estimated Subgraph")
```





## Simulation 3: Highish Dimensional, Cox Model



```{r simulation_params3,echo=FALSE,eval=TRUE}
sim_params$family <- "cox"
dat <- SimulateData(sim_params)
rmean <- max(as.matrix(dat$y)[,1])
```

Same setup as before but with a logistic model linking y and the snps and gene sets. We compute the effects on mean survival restricted to `r round(rmean)`.

```{r plot_make3,echo=FALSE,eval=TRUE}
path_sub <- FindSubgraph(sim_params)
gR <- MakeGraphNELObject(FindSubgraph(sim_params),
                         sim_params$xx_direct[rownames(sim_params$path) %in% rownames(path_sub)],
                         sim_params$mm_direct[colnames(sim_params$path) %in% colnames(path_sub)])
attrs <- list()
attrs$edge <- list()
attrs$edge$fontsize <- 12
edgeAttrs <- MakeedgeAttrs(sim_params$path_model,sim_params$xx_direct,sim_params$mm_direct)
```

```{r compute_estimate3,echo=FALSE,eval=TRUE}
fit <- ComputePath(dat,reg=TRUE)
```

```{r plot_subgraph3,echo=FALSE,eval=TRUE,fig.height=6,fig.width=12}
path_sub <- FindSubgraph(fit)
fit_plot <- fit
fit_plot$path_model <- path_sub
fit_plot$xx_direct <- fit_plot$xx_direct[rownames(fit_plot$path_model)]
fit_plot$mm_direct <- fit_plot$mm_direct[colnames(fit_plot$path_model)]

gR2 <- MakeGraphNELObject(fit_plot$path_model,fit_plot$xx_direct,fit_plot$mm_direct)
attrs2 <- list()
attrs2$edge <- list()
attrs2$edge$fontsize <- 12
edgeAttrs2 <- MakeedgeAttrs(fit_plot$path_model,fit_plot$xx_direct,fit_plot$mm_direct)
par(mfcol=c(1,2))
plot(gR,edgeAttrs=edgeAttrs,attrs=attrs,main="True Subgraph")
plot(gR2,edgeAttrs=edgeAttrs2,attrs=attrs2,main="Estimated Subgraph")
```



```{r xx_effects_decomp3, echo=FALSE,eval=TRUE}
direct <- ComputeEffectxx(dat,fit,"direct",rmean=rmean)
indirect <- ComputeEffectxx(dat,fit,"indirect",rmean=rmean)
total <- ComputeEffectxx(dat,fit,"total",rmean=rmean)
out <- cbind(direct,indirect,total)
rownames(out) <- names(fit$xx_direct)
colnames(out) <- c("direct","indirect","total")
## only output xxs with either direct or indirect effects
## no direct effect
no_direct <- fit$xx_direct==0
## no indirect effect if not connected to any mm which connects to y
no_indirect <- rowSums(t(t(fit$path_model) * fit$mm_direct))==0
out <- out[!(no_direct & no_indirect),]
```



<div class = "row">
<div class="column-left">
```{r kable_table2,echo=FALSE}
kable(out,digits=2) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```
</div>
<div class="column-right">
</div>
</div>