---
title: "Introduction to `mediateR`"
output: 
  rmarkdown::html_vignette:
    number_sections: yes
    toc: yes
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
set.seed(1234)
library(mediateR)
library(ggplot2)
library(ggfortify)
```


# Data Structure

The code contains functions to simulate data for mediation analysis.

```{r simulate}
params <- SimpleSim(n=10,family="gaussian")
dat <- SimulateData(params)
dat
```


The `dat` list with named elements (`r names(dat)`) is used as an argument for functions which estimate the direct and indirect effects. One can manually create a list with these elements (such as when fitting models to new data) or rely on the simulation function `SimulateData` to create the structure. The argument to `SimulateData` has parameters controlling the sample size, coefficient estimates, number of mediators, etc.

# Data EDA

We simulate a decent sized data set for mediation analysis:

```{r simulate_big}
params <- SimpleSim(n=1000,family="gaussian")
dat <- SimulateData(params)
dim(dat$xx)
dim(dat$mm)
```

The goal is to assess the causal impact of `dat$xx` on `dat$y` and quantify how much of this effect passes through `dat$mm`. For some initial evidence, see that $x$ is correlated with both $y$ and $m$ and $m$ and $y$ are correlated:

```{r plots, fig.width=7}
par(mfcol=c(1,3),mar=c(5,4,1,1))
boxplot(dat$y~dat$xx[,1],xlab="x",ylab="y")
boxplot(dat$mm[,1]~dat$xx[,1],xlab="x",ylab="m")
plot(dat$mm,dat$y,xlab="m",ylab="y")
```


# Fit Mediation Model

In linear models, the direct effect is the path coefficient $x \rightarrow m$ and indirect effect is the product of path coefficients $x\rightarrow m$ and $m \rightarrow y$. We compute the effects 3 ways.

1. Exact path coefficients from simulation.
2. Estimated path coefficients from fitting regressions.
3. Numerical integration described in Long et al. 2019

Method 3 is not necessary in linear models (method 2 is easier), but it generalizes to non--linear models where method 2 is no longer applicable. Method 1 is only possible because this is a simulation where we know ground truth.

## Method 1

```{r exact_mediation_effect}
ComputeEffectsLinear(params)
```

## Method 2

```{r path_ests_mediation_effect}
fit <- ComputePath(dat)
ComputeEffectsLinear(fit)
```


## Method 3

```{r numeric_integration}
effect <- c(ComputeEffectxx(dat,fit,"direct"),
            ComputeEffectxx(dat,fit,"indirect"),
            ComputeEffectxx(dat,fit,"total"))
effect
```


# Survival Outcomes

We can also compute effects for survival models.

```{r survival_sim}
params <- SimpleSim(n=1000,family="cox")
dat <- SimulateData(params)
class(dat$y) ## response is Surv object
```


Survival curves suggest very strong effect of `dat$xx` on `dat$y`:

```{r survival_sim_graph}
fit <- survival::survfit(dat$y~as.factor(dat$xx[,1]))
autoplot(fit,xlab="Survival Time (months)",ylab="Survival")
```

Compute mediation effect point estimates:

```{r numeric_integration_surv}
rmean <- 10
fit <- ComputePath(dat)
effect <- c("direct"=ComputeEffectxx(dat,fit,"direct",rmean=rmean),
            "indirect"=ComputeEffectxx(dat,fit,"indirect",rmean=rmean),
            "total"=ComputeEffectxx(dat,fit,"total",rmean=rmean))
effect
```

Appears to be strong negative causal effect with most of this effect mediated by `dat$mm`.

## Bootstrap Sampling to Assess Uncertainty

```{r bootstrap}
B <- 500
boot_effect <- matrix(NA_real_,nrow=B,ncol=2)
for(ii in 1:B){
  ix <- sample(1:params$n,replace=TRUE)
  dat_boot <- mediateR:::SubsetDat(dat,ix)
  fit_boot <- ComputePath(dat_boot)
  boot_effect[ii,1] <- ComputeEffectxx(dat_boot,fit_boot,"direct",rmean=rmean)
  boot_effect[ii,2] <- ComputeEffectxx(dat_boot,fit_boot,"indirect",rmean=rmean)
}
```

Plot bootstrap samples:

```{r bootstrap_plot}
par(mar=c(5,5,1,1))
plot(boot_effect[,1],boot_effect[,2],
     xlim=range(c(0,boot_effect[,1])),ylim=range(c(0,boot_effect[,2])),
     xlab="Direct Effect (months)",ylab="Indirect Effect (months)")
points(effect[1],effect[2],col='red',pch=19)
points(0,0,col='blue',pch=19)
abline(h=0)
abline(v=0)
```

Conclusion: `dat$xx` has a strong negative causal impact on survival with most of this effect passing through the mediator `dat$mm`.




