---
title: "Multiple Mediation Analysis with `mediateR`"
author: James P. Long, PhD. and Licai Huang
output: 
  rmarkdown::html_vignette:
    number_sections: yes
    toc: yes
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Multiple Mediators}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F,message = F)
options(knitr.kable.NA = '')
options(scipen = 999)
```


```{r load_packages}
library(graph)
library(kableExtra)
library(mediateR)
set.seed(280920181)
```
\newcommand{\Var}{\text{Var}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\argmin}[1]{\underset{#1}{\operatorname{argmin}}\text{ }}
\newcommand{\argmax}[1]{\underset{#1}{\operatorname{argmax}}}
\newcommand{\ind}[1]{\mathbf{1}_{#1}}
\newcommand{\todo}[1]{\color{red}{\text{TODO: #1}}}


# Data Structure

The code contains functions to simulate data for mediation analysis.

```{r simulate}
params <- mediateR:::QuickSim(5,3,2,"gaussian")
dat <- SimulateData(params)
dat
```

The `dat` list with named elements (`r names(dat)`) is used as an argument for functions which estimate the direct and indirect effects.

  - `dat$y` is the response,
  - `dat$mm` is the vector of the mediators, 
  - `dat$xx` is the vector of the exposures, 
  - `dat$co` is the vector of the covariates, and 
  - `dat$family` denotes the type of response.

One can manually create a list with these elements (such as when fitting models to new data) or rely on the simulation function `SimulateData` to create the structure. The argument to `SimulateData` has parameters controlling the sample size, coefficient estimates, number of mediators, etc.


# Mediation Analysis

The goal is to assess the causal impact of `dat$xx` on `dat$y` and quantify how much of this effect passes through `dat$mm`, adjusting for covariates `dat$co`. 


## Linear

We simulate `n` observations from the "True Graph" using independent, binary SNPs drawn from Bernoulli(1/2) and path coefficients given below.

```{r simulation_params,echo=TRUE,eval=TRUE}
n <- 500
sim_params <- mediateR:::QuickSim(n,3,2,"gaussian")
dat <- SimulateData(sim_params)
```

```{r plot_make,echo=F,eval=F}
gR <- mediateR:::MakeGraphNELObject(sim_params$path,sim_params$xx_direct,sim_params$mm_direct)
attrs <- list()
attrs$edge <- list()
attrs$edge$fontsize <- 12
edgeAttrs <- mediateR:::MakeedgeAttrs(sim_params$path_model,sim_params$xx_direct,sim_params$mm_direct)
```


### Model fitting

`ComputePath` is used to fit the mediation analysis framework for various response, where `dat` are the datalist with the format mentioned above, `mmn=F` assuming independent mediators and `reg=T` is used for high-dimensional mediators (default is false) 


```{r compute_estimate,echo=TRUE,eval=TRUE}
fit <- ComputePath(dat, mmn = F, reg = F)
eff_est <- ComputeEffectsLinear(fit)
```

In linear models, the direct effect is the path coefficient $x \rightarrow m$ (`xx_direct`),  and indirect effect is the product of path coefficients $x\rightarrow m$ (`path_model`), and $m \rightarrow y$ (`mm_direct`) .

```{r plot_subgraph,echo=F,eval=F,fig.height=6,fig.width=12}
gR2 <- mediateR:::MakeGraphNELObject(fit$path_model,fit$xx_direct,fit$mm_direct)
attrs2 <- list()
attrs2$edge <- list()
attrs2$edge$fontsize <- 12
edgeAttrs2 <- mediateR:::MakeedgeAttrs(fit$path_model,fit$xx_direct,fit$mm_direct)
par(mfcol=c(1,2))
plot(gR,edgeAttrs=edgeAttrs,attrs=attrs,main="True Graph")
plot(gR2,edgeAttrs=edgeAttrs2,attrs=attrs2,main="Estimated Graph")
```


```{r merge_true_est,echo=TRUE,eval=TRUE}
eff <- ComputeEffectsLinear(sim_params)
eff_comb <- matrix(0,nrow(eff),ncol=2*ncol(eff))
for(ii in 1:ncol(eff)){
  eff_comb[,2*ii-1] <- eff_est[,ii]
  eff_comb[,2*ii] <- eff[,ii]
}
for(ii in (ncol(dat$xx)+1):(ncol(dat$xx)+ncol(dat$mm))){
  eff_comb[ii,3:ncol(eff_comb)] <- NA
}
colnames(eff_comb) <- rep(c("est","true"),ncol(eff))
rownames(eff_comb) <- rownames(eff)
```

Table below summarizes the Indirect, Direct, and Total effects. 

```{r kable_table,echo=TRUE}
kable(eff_comb,digits=2) %>%
  kable_styling(bootstrap_options = "striped", full_width = F) %>%
  add_header_above(c(" ", "direct" = 2, "indirect" = 2, "total" = 2))
```

### Bootstrap 

We propose computing confidence intervals and hypothesis tests using bootstrap sampling quantiles.

```{r bootstrap,echo=TRUE,eval=TRUE}
B <- 1000
eff_est_boot <- vector("list",length=B)
for(ii in 1:B){
  ix <- sample(1:sim_params$n,replace=TRUE)
  dat_sub <- mediateR:::SubsetDat(dat,ix)
  fit <- ComputePath(dat_sub)
  eff_est_boot[[ii]] <- ComputeEffectsLinear(fit)
}

total_xx1 <- vapply(eff_est_boot,function(x){x[1,3]},c(0))
L95 = round(quantile(total_xx1, 0.025),2)
U95 = round(quantile(total_xx1, 1 - 0.025),2)
```
Figure below shows the histgram of total effects of Bootstrap samples for SNP1. The 95\% CI of the total effect for SNP1 is (`r L95`, `r U95`)  

```{r bootstrap_plot,echo=TRUE,fig.height=4.5,fig.width=4.5}

hist(total_xx1,main="Bootstrap Samples SNP1 Total Effect",xlab="Estimate")
abline(v=eff[1,3],lwd=3,col='red')
abline(v=eff_est[1,3],lwd=3,col='blue')
legend("topright",c("Truth","Point Estimate"),col=c("red","blue"),lwd=2)
```



## Logistic

Same as simulation 1 but with binomial link function connecting y with snps and gene sets.

```{r simulation_params2,echo=TRUE,eval=TRUE}
sim_params$family <- "binomial"
dat <- SimulateData(sim_params)
```

```{r plot_make2,echo=F,eval=F}
gR <- mediateR:::MakeGraphNELObject(sim_params$path,sim_params$xx_direct,sim_params$mm_direct)
attrs <- list()
attrs$edge <- list()
attrs$edge$fontsize <- 12
edgeAttrs <- mediateR:::MakeedgeAttrs(sim_params$path_model,sim_params$xx_direct,sim_params$mm_direct)
```

### Model Fitting 

```{r compute_estimate2,echo=TRUE,eval=TRUE}
fit <- ComputePath(dat)
```

```{r plot_subgraph2,echo=F,eval=F,fig.height=6,fig.width=12}
gR2 <- mediateR:::MakeGraphNELObject(fit$path_model,fit$xx_direct,fit$mm_direct)
attrs2 <- list()
attrs2$edge <- list()
attrs2$edge$fontsize <- 12
edgeAttrs2 <- mediateR:::MakeedgeAttrs(fit$path_model,fit$xx_direct,fit$mm_direct)
par(mfcol=c(1,2))
plot(gR,edgeAttrs=edgeAttrs,attrs=attrs,main="True Graph")
plot(gR2,edgeAttrs=edgeAttrs2,attrs=attrs2,main="Estimated Graph")
```
Given the simulation setting, we can approximate the true effects by approximating integral with large sample size by setting `n` to million observations.

```{r xx_effects_decomp2_t, echo=TRUE,eval=TRUE}
sim_params2 <- sim_params
sim_params2$n <- 1e6 
dat2 <- SimulateData(sim_params2)
direct_t <- ComputeEffectxx(dat2,sim_params,"direct")
indirect_t <- ComputeEffectxx(dat2,sim_params,"indirect")
total_t <- ComputeEffectxx(dat2,sim_params,"total")
```

`ComputeEffectxx` provides estimation of direct, indirect and total effect.

```{r  xx_effects_decomp2, echo=TRUE,eval=TRUE}
direct <- ComputeEffectxx(dat,fit,"direct")
indirect <- ComputeEffectxx(dat,fit,"indirect")
total <- ComputeEffectxx(dat,fit,"total")
eff_comb <- cbind(direct,direct_t,indirect,indirect_t,total,total_t)
rownames(eff_comb) <- names(fit$xx_direct)
colnames(eff_comb) <- c("est","true","est","true","est","true")
xx1total_sample <- eff_comb[1,5]
xx1total_true <- eff_comb[1,6]
```

Table below summarizes the Indirect, Direct, and Total effects. 

```{r kable_table2,echo=TRUE}
kable(eff_comb,digits=2) %>%
  kable_styling(bootstrap_options = "striped", full_width = F) %>%
  add_header_above(c(" ", "direct" = 2, "indirect" = 2, "total" = 2))
```

### Bootstrap 

```{r bootstrap2,echo=TRUE,eval=TRUE}
B <- 1000
total_xx1 <- rep(0,length=B)
for(ii in 1:B){
  ix <- sample(1:sim_params$n,replace=TRUE)
  dat_boot <- mediateR:::SubsetDat(dat,ix)
  fit_boot <- ComputePath(dat_boot)
  total_xx1[ii] <- ComputeEffectxx(dat_boot,fit_boot,"total")[1]
}
L95 = round(quantile(total_xx1, 0.025),2)
U95 = round(quantile(total_xx1, 1 - 0.025),2)
```

Figure below shows the histgram of total effects of Bootstrap samples for SNP1. The 95\% CI of the total effect for SNP1 is (`r L95`, `r U95`)  


```{r bootstrap_plot2,echo=TRUE,fig.height=4.5,fig.width=4.5}
hist(total_xx1,main="Bootstrap Samples SNP1 Total Effect",xlab="Estimate")
abline(v=xx1total_sample,lwd=3,col='blue')
abline(v=xx1total_true,lwd=3,col='red')
legend("topright",c("Point Estimate","True"),col=c("blue","red"),lwd=2)
```


## Survival

Same as simulation 1 and 3 but with exponentially distributed survival times. The rate is $e^{\beta^Tx}$ where $x$ are the covariates. This model can be fit with Cox Proportional Hazards.

```{r simulation_params3,echo=TRUE,eval=TRUE}
sim_params$family <- "cox"
dat <- SimulateData(sim_params)
rmean <- max(as.matrix(dat$y)[,1])
```

### Model Fitting 

Same as simulation 1 and 3 but with exponentially distributed survival times. The rate is $e^{\beta^Tx}$ where $x$ are the covariates. This model can be fit with Cox Proportional Hazards. We compute the effects on mean survival restricted to `r round(rmean)`.

```{r plot_make3,echo=F,eval=F}
gR <- mediateR:::MakeGraphNELObject(sim_params$path,sim_params$xx_direct,sim_params$mm_direct)
attrs <- list()
attrs$edge <- list()
attrs$edge$fontsize <- 12
edgeAttrs <- mediateR:::MakeedgeAttrs(sim_params$path_model,sim_params$xx_direct,sim_params$mm_direct)
```

```{r compute_estimate3,echo=TRUE,eval=TRUE}
fit <- ComputePath(dat)
```

```{r plot_subgraph3,echo=F,eval=F,fig.height=6,fig.width=12}
gR2 <- mediateR:::MakeGraphNELObject(fit$path_model,fit$xx_direct,fit$mm_direct)
attrs2 <- list()
attrs2$edge <- list()
attrs2$edge$fontsize <- 12
edgeAttrs2 <- mediateR:::MakeedgeAttrs(fit$path_model,fit$xx_direct,fit$mm_direct)
par(mfcol=c(1,2))
plot(gR,edgeAttrs=edgeAttrs,attrs=attrs,main="True Graph")
plot(gR2,edgeAttrs=edgeAttrs2,attrs=attrs2,main="Estimated Graph")
```
Given the simulation setting, we approximate true effects by approximating integral with large sample size use the maximum observed time as the restricted mean time.

```{r xx_effects_decomp3_t, echo=TRUE,eval=TRUE}

sim_params2 <- sim_params
sim_params2$n <- 1e4 # use many observations to approximate effects
dat2 <- SimulateData(sim_params2)
fit2 <- ComputePath(dat2) 

direct_t <- ComputeEffectxx(dat2,fit2,"direct",rmean=rmean)
indirect_t <- ComputeEffectxx(dat2,fit2,"indirect",rmean=rmean)
total_t <- ComputeEffectxx(dat2,fit2,"total",rmean=rmean)
```

Similarly, `ComputeEffectxx` provides estimation of direct, indirect and total effect. `rmean`, the mean survival restricted to `r round(rmean)`, needs to be provided to calculate the effect.

```{r xx_effects_decomp3, echo=TRUE,eval=TRUE}
direct <- ComputeEffectxx(dat,fit,"direct",rmean=rmean)
indirect <- ComputeEffectxx(dat,fit,"indirect",rmean=rmean)
total <- ComputeEffectxx(dat,fit,"total",rmean=rmean)
eff_comb <- cbind(direct,direct_t,indirect,indirect_t,total,total_t)
rownames(eff_comb) <- names(fit$xx_direct)
colnames(eff_comb) <- c("est","true","est","true","est","true")
xx1total_sample <- eff_comb[1,5]
xx1total_true <- eff_comb[1,6]
```

Table below summarizes the Indirect, Direct, and Total effects. 

```{r kable_table3,echo=TRUE}
kable(eff_comb,digits=2) %>%
  kable_styling(bootstrap_options = "striped", full_width = F) %>%
  add_header_above(c(" ", "direct" = 2, "indirect" = 2, "total" = 2))
```


### Bootstrap 

```{r bootstrap3,echo=TRUE,eval=TRUE}
B <- 1000
total_xx1 <- rep(0,length=B)
for(ii in 1:B){
  ix <- sample(1:sim_params$n,replace=TRUE)
  dat_boot <- mediateR:::SubsetDat(dat,ix)
  fit_boot <- ComputePath(dat_boot)
  total_xx1[ii] <- ComputeEffectxx(dat_boot,fit_boot,"total",rmean=rmean)[1]
}
L95 = round(quantile(total_xx1, 0.025),2)
U95 = round(quantile(total_xx1, 1 - 0.025),2)
```

Figure below shows the histgram of total effects of Bootstrap samples for SNP1. The 95\% CI of the total effect for SNP1 is (`r L95`, `r U95`)  


```{r bootstrap_plot3,echo=TRUE,fig.height=4.5,fig.width=4.5}
hist(total_xx1,main="Bootstrap Samples SNP1 Total Effect",xlab="Estimate")
abline(v=xx1total_sample,lwd=3,col='blue')
abline(v=xx1total_true,lwd=3,col='red')
legend("topright",c("Point Estimate","True"),col=c("blue","red"),lwd=2)
```

### Estimation of mediator-mediator relations. 

Partial correlation between mediators adjusting for exposures and covariates

```{r compute_estimate4,echo=TRUE,eval=TRUE}
sim_params <- mediateR:::QuickSim(n,4,5,"gaussian")
sim_params$family <- "cox"
dat <- SimulateData(sim_params)

dat <- SimulateData(sim_params)
rmean <- max(as.matrix(dat$y)[,1])

fit <- ComputePath(dat, mmn = T)
par1 = cov2cor(fit$cov_mm)
colnames(par1) = rownames(par1) = names(fit$mm_direct)
kable(par1,digits=2) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```

Partial correlation between mediators adjusting for other mediators, exposures and covariates

```{r icov,echo=TRUE,eval=TRUE}
par2 = fit$ivcor_mm
colnames(par2) = rownames(par2) = names(fit$mm_direct)
kable(par2,digits=2) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)

```


Here we show how to compute confidence intervals and hypothesis tests using bootstrap sampling quantiles.

```{r icov_bootstrap,echo=TRUE,eval=TRUE}
B <- 100
corrmm <- array(0,dim = c(B, ncol(dat$mm),ncol(dat$mm)))
for(ii in 1:B){
  ix <- sample(1:sim_params$n,replace=TRUE)
  dat_boot <- mediateR:::SubsetDat(dat,ix)
  fit_boot <- ComputePath(dat_boot, mmn = T)
  corrmm[ii,,] <- fit_boot$ivcor_mm
}

L95 = round(quantile(corrmm[,1,2], 0.025),2)
U95 = round(quantile(corrmm[,1,2], 1 - 0.025),2)

```

The 95\% CI of partial correlation adjusting for other mediators, exposures and covariates between `m1` and `m2` is (`r L95`, `r U95`). 0 is included in this interval suggesting that the partial correlation between `m1` and `m2` is not significant. 
